---
number-sections: true
number-offset: [6, 0]
---

# Long Short-Term Memory (LSTM) Networks {#sec-lstm-networks}

## Overview

The [previous chapter](recurrent_neural_networks.qmd) showed the central weakness of a plain recurrent neural network (RNN): the hidden state is recursively updated, but the gradients used for learning can vanish as they are propagated backward through many time steps. Long Short-Term Memory (LSTM) networks were designed to address precisely this problem.

LSTMs add a dedicated **cell state** $C_t$, a memory path through time whose attenuation the network itself controls through **gates** — small neural-network components that decide what information to retain, what new information to add, and what part of the internal memory to expose as the hidden state.

For econometricians, LSTMs are useful when the relevant predictive state may evolve over time and depend on more than the most recent observations. Examples include asset pricing with time-varying macroeconomic states, macroeconomic forecasting with long predictor histories, and volatility or risk forecasting with persistent regimes.

For optional visual intuition, [this blog post](https://colah.github.io/posts/2015-08-Understanding-LSTMs/) gives a detailed informal discussion of LSTMs.

## Roadmap

1. We first introduce the LSTM cell state, hidden state, and gates.
2. We then work through the LSTM update equations and notation.
3. Next, we explain why the cell state improves gradient flow relative to a plain RNN.
4. We discuss when the additional complexity of an LSTM is useful in econometric forecasting, and what changes — and what does not — when training one.
5. We present a research application in asset pricing.
6. We close with a look ahead from recurrent states to attention — a construction that lets a model consult earlier positions in a sequence directly instead of compressing the entire past into one state — plus key takeaways, common pitfalls, and manual forward-pass and gradient-flow exercises.

## LSTM Architecture and Notation

LSTM networks, introduced by @HochreiterSchmidhuber1997, extend plain RNNs by adding a gating mechanism that controls long-term information flow; the forget gate, central to the modern form of the cell, was added later by @GersSchmidhuberCummins2000. The notation below lists the quantities that the LSTM tracks at each time step.

At each time step $t$, an LSTM cell maintains the following quantities, where $H$ is the number of hidden units and $p$ the input dimension. To keep the gate equations light, we drop the boldface used for vectors in the [RNN chapter](recurrent_neural_networks.qmd): $x_t$, $h_t$, $C_t$, and the gates below are all vectors of the stated dimensions, and the weight matrices $W_f, W_i, W_c, W_o$ are likewise set without boldface.

- **Input**: $x_t \in \mathbb{R}^p$, the current input vector
- **Hidden state**: $h_t \in \mathbb{R}^H$, the output of the LSTM cell, similar to RNN hidden states
- **Cell state**: $C_t \in \mathbb{R}^H$, the internal memory of the LSTM cell

The LSTM uses three gates and the candidate values (all in $\mathbb{R}^H$):

- **Forget gate**: $f_t$, controls which information from the previous cell state to retain or discard
- **Input gate**: $i_t$, controls what new information to store in the cell state
- **Candidate values**: $\tilde{C}_t$, new candidate information that could be added to the cell state
- **Output gate**: $o_t$, controls what parts of the cell state to output as the hidden state

**Update Equations and Weight Matrices**

Let $\sigma$ denote the sigmoid function and $\tanh$ the hyperbolic tangent (mapping to $(-1,1)$). In each equation below, $[h_{t-1}, x_t]$ stacks the previous hidden state and the current input into a single vector in $\mathbb{R}^{H+p}$, and $\cdot$ denotes matrix-vector multiplication (the same symbol is also used later for ordinary scalar multiplication). Finally, $\odot$ denotes the element-wise (Hadamard) product. The LSTM cell state and hidden state are updated according to:

$$
\begin{aligned}
f_t &= \sigma(W_f \cdot [h_{t-1}, x_t] + b_f) \quad \text{(forget gate)} \\
i_t &= \sigma(W_i \cdot [h_{t-1}, x_t] + b_i) \quad \text{(input gate)} \\
\tilde{C}_t &= \tanh(W_c \cdot [h_{t-1}, x_t] + b_c) \quad \text{(candidate values)} \\
C_t &= f_t \odot C_{t-1} + i_t \odot \tilde{C}_t \quad \text{(cell state update)} \\
o_t &= \sigma(W_o \cdot [h_{t-1}, x_t] + b_o) \quad \text{(output gate)} \\
h_t &= o_t \odot \tanh(C_t) \quad \text{(hidden state)}
\end{aligned}
$$

Each gate and the candidate values are computed using their own weight matrices and bias vectors.
Since the input to each is the concatenation $[h_{t-1}, x_t] \in \mathbb{R}^{H+p}$, the dimensions are:

- $W_f, W_i, W_c, W_o \in \mathbb{R}^{H \times (H+p)}$: weight matrices
- $b_f, b_i, b_c, b_o \in \mathbb{R}^H$: bias vectors

Equivalently, partitioning each weight matrix into its state and input blocks — $W_f = [W_{f,h} \; W_{f,x}]$ with $W_{f,h} \in \mathbb{R}^{H \times H}$ and $W_{f,x} \in \mathbb{R}^{H \times p}$ — gives $W_f \cdot [h_{t-1}, x_t] = W_{f,h}\, h_{t-1} + W_{f,x}\, x_t$: the concatenated form used here and the two-matrix form used for the plain RNN in the [previous chapter](recurrent_neural_networks.qmd#sec-rnn-architecture) describe the same model.

**The Gating Mechanism**

A "gate" is a mechanism to selectively let information pass. In LSTMs, this is achieved by combining a sigmoid activation function with element-wise multiplication. The sigmoid function squashes any input to a value between 0 and 1. This output vector then acts as a gatekeeper:

-   A gate value close to **0** means "let almost nothing through" (a nearly closed gate).
-   A gate value close to **1** means "let almost everything through" (a nearly open gate).
-   Intermediate values pass information through partially. (The sigmoid approaches 0 and 1 without attaining them — a point we return to below.)


@fig-lstm-cell traces how these pieces fit together within a single time step. When reading it, note in particular how short the top path from $C_{t-1}$ to $C_t$ is compared with everything else in the cell.

```{r}
#| label: fig-lstm-cell
#| fig-cap: "Flow of information through one LSTM cell. The previous hidden state $h_{t-1}$ and the current input $x_t$ enter at the lower left and feed four parallel blocks: three sigmoid gates ($\\sigma$, yellow) producing the forget gate $f_t$, the input gate $i_t$, and the output gate $o_t$, and one tanh block (green) producing the candidate $\\tilde{C}_t$. Circles marked $\\odot$ and $+$ apply elementwise multiplication and addition. Along the top, the previous cell state $C_{t-1}$ is scaled by $f_t$, augmented by $i_t \\odot \\tilde{C}_t$, and passed on as $C_t$ — the long-term memory highway. The new hidden state $h_t$ (lower right) is the output gate applied to $\\tanh(C_t)$."
#| fig-align: center
#| echo: false
#| message: false
#| warning: false

required_r_packages <- c("tinytex", "pdftools", "png")
missing_r_packages <- required_r_packages[
  !vapply(required_r_packages, requireNamespace, logical(1), quietly = TRUE)
]
if (length(missing_r_packages) > 0) {
  stop("Install the required R packages before rendering: ",
       paste(missing_r_packages, collapse = ", "))
}

# Build PDF then PNG
if (!file.exists("figures/lstm_cells.pdf") ||
    file.mtime("figures/lstm_cells.pdf") < file.mtime("figures/lstm_cells.tex")) {
  invisible(tinytex::latexmk("figures/lstm_cells.tex", engine = "pdflatex", clean = TRUE))
}
if (!file.exists("figures/lstm_cells.png") ||
    file.mtime("figures/lstm_cells.png") < file.mtime("figures/lstm_cells.pdf")) {
  bmp <- pdftools::pdf_render_page("figures/lstm_cells.pdf", page = 1, dpi = 220)
  png::writePNG(bmp, "figures/lstm_cells.png")
}

paths <- if (knitr::is_html_output()) {
  c("figures/lstm_cells.png")
} else {
  c("figures/lstm_cells.pdf")
}
knitr::include_graphics(paths)
```

@fig-lstm-cell makes the division of labor visible. The top path — from $C_{t-1}$ through one elementwise multiplication and one addition to $C_t$ — is short and nearly linear: this is the route along which information, and later gradient signal, can travel across many periods, and the next section analyzes exactly this path. Everything else in the cell exists to regulate it: the gates decide how much of the old state survives ($f_t$), how much new information enters ($i_t \odot \tilde{C}_t$), and how much of the internal memory is exposed to the rest of the network as $h_t$.

## Gradient Flow in LSTMs

LSTMs address the vanishing-gradient problem of plain RNNs through the design of the cell state. Its update rule is
$$
C_t = f_t \odot C_{t-1} + i_t \odot \tilde{C}_t
$$

The update has an additive structure. The old state enters through $f_t\odot C_{t-1}$, while new information enters through $i_t\odot\tilde C_t$. To isolate the direct cell-state path, hold the gates $f_t,i_t,o_t$ and the candidate $\tilde C_t$ fixed. Because the update acts elementwise, its Jacobian is

$$
\frac{\partial C_t}{\partial C_{t-1}}\bigg|_{\text{gates fixed}} = \operatorname{diag}(f_t),
$$
so for coordinate $j$,
$$
\frac{\partial C_{t,j}}{\partial C_{t-1,j}}\bigg|_{\text{gates fixed}} = f_{t,j},
$$
and chaining over $T-t$ steps gives
$$
\frac{\partial C_{T,j}}{\partial C_{t,j}}\bigg|_{\text{gates fixed}} = \prod_{k=t+1}^T f_{k,j}.
$$

The structural contrast with a plain RNN is visible by placing the two paths side by side:
$$
\begin{aligned}
\text{Plain RNN:} \quad &\frac{\partial h_T}{\partial h_t} = \prod_{k=t+1}^T \frac{\partial h_k}{\partial h_{k-1}} = \prod_{k=t+1}^T \operatorname{diag}\!\bigl(\tanh'(W_{hh} h_{k-1} + W_{xh} x_k + b_h)\bigr)\, W_{hh} \\
\text{LSTM (cell-state path, gates fixed, coordinate } j\text{):} \quad &\frac{\partial C_{T,j}}{\partial C_{t,j}} = \prod_{k=t+1}^T f_{k,j}
\end{aligned}
$$

Every plain-RNN factor contains both a recurrent transformation and an activation derivative. The direct LSTM cell-state path instead bypasses $\tanh$, and its attenuation is controlled by the input-dependent forget gate $f_t=\sigma(W_f\cdot[h_{t-1},x_t]+b_f)$ rather than by $\|W_{hh}\|$ — a statement about this one path only, since the gates are themselves functions of $W_f$ and the full parameter gradient still depends on the weights through them. A value near zero removes old information; a value near one retains it; and different coordinates can operate at different rates. If the relevant forget gates remain close to one, then
$$
\frac{\partial C_{T,j}}{\partial C_{t,j}} = \prod_{k=t+1}^T f_{k,j} \approx 1^{T-t} = 1
$$

The near-unit forget-gate product means that gradients can travel many steps with little attenuation. The decay is never switched off entirely: each finite-horizon product lies strictly below one; it shrinks geometrically when the gates stay uniformly below some $\bar{f}<1$ and, when the gate is constant, decays at rate $f_j^{\,T-t}$. The LSTM can therefore learn the rate of retention separately by coordinate and context. In a plain RNN, the same recurrent transformation helps determine both the state dynamics and the gradient path, while every step also contributes an activation derivative.

Unlike plain RNNs, where gradient flow is shaped by repeated multiplication with the recurrent weight matrix, LSTMs *learn* when to let gradients flow through the forget gates. This learned control is the mechanism by which an LSTM can *learn* long-run dependence more reliably — recall from [Exercise 6.1](recurrent_neural_networks.qmd#rnn-exercise-6-1) that the plain RNN's problem is one of training signal, not of what the architecture can represent.

The gates-fixed expression describes only the direct cell-state path. In a full backward pass, the gates themselves depend on $h_{k-1}=o_{k-1}\odot\tanh(C_{k-1})$. Parameter gradients therefore also travel through the hidden state and gate-producing networks, where products of sigmoid and tanh derivatives can attenuate as in a plain RNN. The cell state supplies a low-attenuation route for one component of the backward pass; it does not eliminate all vanishing gradients. Whether a fitted LSTM learns useful long-run dependence remains an empirical question to be judged by validation at deployment-relevant horizons.

## LSTM vs. Plain RNN

The architectural and gradient-flow comparisons summarize the mechanisms introduced by @HochreiterSchmidhuber1997 and @GersSchmidhuberCummins2000. The parameter counts follow directly from the displayed update equations. The computational comparison counts affine blocks per time step; it is not a claim that an LSTM takes exactly four times as long in every implementation.

| Aspect                | Plain RNN                                                              | LSTM                                                                    |
|-----------------------|------------------------------------------------------------------------|-------------------------------------------------------------------------|
| **Memory capability** | Can represent long-run dependence, but its long-gap training signal may vanish | Adds a controlled cell-state path that can retain long-run information more reliably |
| **Gradient flow**     | Products of recurrent weight matrices and activation derivatives       | Additive cell-state path with products of forget-gate values            |
| **Vanishing gradients** | Highly susceptible along the recurrent-state path                     | Mitigated along the cell-state path, but not eliminated throughout the network |
| **Parameters (recurrent cell)** | $H(H+p) + H$                                                 | $4H(H+p) + 4H$ (four weight blocks, approximately four times the RNN cell) |
| **Computational cost**  | One recurrent affine block per time step                              | Four gate or candidate affine blocks per time step; wall-clock cost is implementation-dependent |

**When to use LSTM:**

-   Forecasting tasks where information from distant observations plausibly matters, such as macroeconomic state dynamics or asset-pricing applications with persistent risk states.
-   Settings where validation performance justifies the extra parameters and computational cost.

**When a plain RNN might suffice:**

-   Problems with very short sequences where long-term memory is not needed.
-   When computational resources are extremely limited.

:::{.callout-note title="Question for Reflection"}

Suppose an LSTM is trained on monthly data and the forget gate learns values close to 1 for one hidden dimension and close to 0 for another. Using the gates-fixed coordinatewise gradient expression $\partial C_{T,j} / \partial C_{t,j} = \prod_{k=t+1}^{T} f_{k,j}$, which coordinate attenuates information more slowly? Can the phrase "close to 1" alone justify saying that its gradient is essentially undamped over an arbitrary horizon?
:::

:::{.callout-tip collapse="true" title="Suggested Answer"}

The high-retention coordinate attenuates information more slowly, while the low-retention coordinate's product quickly collapses toward zero. But "close to 1" is not enough to quantify retention without both a numerical gate value and a horizon: for example, $0.99^{30} \approx 0.74$, whereas $0.99^{300} \approx 0.05$. The LSTM can therefore dedicate some coordinates to slower-moving state and others to local dynamics, but even a high-retention coordinate is not undamped over every horizon.
:::

## Training an LSTM

The core estimation framework remains the same when moving from a plain RNN to an LSTM. Training minimizes a loss $L$ over windows cut from the series, gradients are computed on the unrolled cell by backpropagation through time, and the fixed-window and truncated-backpropagation choices of the [previous chapter](recurrent_neural_networks.qmd#sec-rnn-windows) still apply. Under fixed-window training — the scheme used in the [empirical chapter](nn_example.qmd) — both the hidden state and the cell state are reset to $h_0=0$ and $C_0=0$ at the start of every window, so the forward state horizon and the backward gradient horizon both end at the window boundary. Under the state-carrying variant of truncated backpropagation (called *stateful* training in Keras), the final $h$ and $C$ of one segment are passed to the next contiguous segment while the gradient is detached at the boundary, so the forward state can carry information from before the current segment even though the gradient does not reach back that far. What changes relative to a plain RNN, under either scheme, is what flows through the backward pass: alongside the hidden-state path, the cell-state path contributes the gate-controlled products analyzed above.

Two practical differences deserve mention. First, the parameter count roughly quadruples relative to a plain RNN cell of the same width ($4H(H+p)+4H$ against $H(H+p)+H$, as in the table above), so overfitting concerns and validation discipline bite correspondingly harder in the small samples typical of macroeconomic applications. Second, gradient clipping remains standard practice: the cell state tames the vanishing problem along one path, but gradients can still explode through the gate-producing networks, and the clipping safeguard from the [RNN training discussion](recurrent_neural_networks.qmd#sec-rnn-training) carries over unchanged. In Keras, finally, the switch is one line — an `LSTM` layer in place of a `SimpleRNN` layer — as the [empirical chapter](nn_example.qmd) shows in its volatility application.

## Research Application: Deep Learning in Asset Pricing

To see how LSTMs are used in current econometric research, consider two closely related lines of work in empirical asset pricing: the return-prediction approach of @GuKellyXiu2020 and the no-arbitrage approach of @ChenPelgerZhu2024, which embeds an LSTM at its core.

**Two research objectives.** Both papers ask whether neural networks can improve on linear specifications of the relation between firm characteristics, macroeconomic information, and stock returns, but they target different objects. @GuKellyXiu2020 predict returns, whereas @ChenPelgerZhu2024 estimate a no-arbitrage **stochastic discount factor (SDF)**, the pricing kernel implied by the absence of arbitrage. Let $\mathcal{I}_{i,t}$ denote the macroeconomic and firm-specific predictors available for stock $i$ at time $t$, and define the conditional-mean target
$$
m_{i,t}=\mathbb{E}[r_{i,t+1}\mid\mathcal{I}_{i,t}].
$$
A linear characteristic model restricts this target to $m_{i,t}=\beta_0+\sum_{k=1}^{K}\beta_k\,\text{characteristic}_{i,t,k}$, whereas a neural network uses a flexible function $m_{i,t}=f_\theta(\mathcal{I}_{i,t})$. The latter permits nonlinearities and interactions that would otherwise have to be chosen manually. The two papers differ in whether that flexibility is disciplined by a prediction loss or by asset-pricing moment restrictions.

**Return prediction with a feed-forward network.** @GuKellyXiu2020 frame the task as a large-scale supervised prediction problem: a panel regression with a flexible functional form.

-   **Inputs ($\mathbf{x}_{i,t}$):** In the paper's baseline, 94 firm-specific characteristics for stock $i$ at time $t$ (e.g., size, value, momentum, profitability), which are further interacted with eight aggregate time-series predictors and combined with industry indicators — over 900 baseline signals in total. The illustration below uses the 94 characteristics alone.
-   **Target ($y_{i,t+1}$):** The realized excess return of stock $i$ in the next month.
-   **Objective:** Train a neural network $f_\theta$ to minimize the mean squared error (MSE) between predicted and realized returns over a large dataset of all U.S. stocks from 1957 to 2016.
    $$ \min_\theta \frac{1}{N \cdot T} \sum_{i,t} (r_{i,t+1} - f_\theta(\mathbf{x}_{i,t}))^2 $$
    The display is schematic: the U.S. stock panel is unbalanced, so the actual average runs over the observed stock-month pairs rather than all $N \cdot T$ combinations.

The trained network is a **return forecaster**: economics enters primarily through the target, predictors, and portfolio evaluation, rather than through a no-arbitrage estimating equation. A natural architecture is a plain feed-forward network for the cross-sectional relationship; @fig-gkx-architecture sketches its shape.

```{mermaid}
%%| label: fig-gkx-architecture
%%| fig-cap: "Feed-forward architecture for the Gu, Kelly, and Xiu (2020)-style return forecaster, simplified to the 94 firm characteristics (the paper's full baseline design adds macro interactions and industry indicators). The characteristics enter an input layer, pass through three fully connected hidden layers of decreasing width (32, 16, and 8 units with rectified linear unit (ReLU) activations), and produce a single linear output: the return forecast."
graph LR
    A["Input Layer<br/>(94 Characteristics)"] --> B["Hidden Layer 1<br/>(32 neurons, ReLU)"]
    B --> C["Hidden Layer 2<br/>(16 neurons, ReLU)"]
    C --> D["Hidden Layer 3<br/>(8 neurons, ReLU)"]
    D --> E["Output Layer<br/>(1 neuron, linear)"]

    style A fill:#e6f2ff,stroke:#333
    style B fill:#fff2e6,stroke:#333
    style C fill:#fff2e6,stroke:#333
    style D fill:#fff2e6,stroke:#333
    style E fill:#ffe6e6,stroke:#333
```

The depth of the network allows it to learn a **hierarchy of features**. For example, the first layer might learn to combine basic accounting ratios into a value signal, and a deeper layer could then learn to model the interaction between this value signal and momentum signals.

**From prediction to a no-arbitrage objective.** @ChenPelgerZhu2024 take a deliberately different route. Rather than forecasting returns, they estimate $M_{t+1}$ directly, and they replace the MSE objective with Euler-equation restrictions that characterize admissible SDFs:
$$
\mathbb{E}\big[M_{t+1}\, R^e_{i,t+1}\, g(I_t, I_{i,t})\big] = 0
\quad \text{for all conditioning functions } g,
$$
where $R^e_{i,t+1}$ is the excess return on stock $i$, $I_t$ and $I_{i,t}$ collect the macroeconomic and firm-specific information available at time $t$, and $g(\cdot)$ provides test assets, or instruments. In generalized method of moments (GMM), parameters are chosen to make sample analogues of model-implied moment restrictions as close to zero as possible. Conditional GMM allows those restrictions to be multiplied by functions of information available at time $t$; here, $g$ creates those functions. Their data are 46 firm characteristics together with 178 macroeconomic time series, for U.S. stocks from 1967 to 2016. The estimation principle is therefore conditional GMM, not least-squares prediction. The authors argue that this distinction carries real weight: in their comparisons, machine-learning models trained on a pure prediction objective can perform *worse* than linear models that impose no-arbitrage.

**A recurrent macro state.** Motivated by time-varying risk premia and a persistent macroeconomic state, @ChenPelgerZhu2024 use an LSTM to process the macroeconomic sequence and compress it into a small hidden state $h_t$. Feeding all 178 raw macro series into the SDF network each month would treat the latest releases as the complete state, whereas $h_t$ is a learned, low-dimensional summary of macroeconomic *history*. This state vector enters the SDF network alongside the firm characteristics.

```{r}
#| label: fig-pelger-architecture
#| fig-cap: "Architecture of the SDF model of Chen, Pelger, and Zhu (2024). The sequence of macroeconomic variables is compressed by an LSTM into a low-dimensional macro state vector $h_t$; this state is concatenated with the firm characteristics and passed through a feed-forward network whose outputs are the SDF portfolio weights $\\omega$."
#| fig-align: center
#| echo: false
#| message: false
#| warning: false

required_r_packages <- c("tinytex", "pdftools", "png")
missing_r_packages <- required_r_packages[
  !vapply(required_r_packages, requireNamespace, logical(1), quietly = TRUE)
]
if (length(missing_r_packages) > 0) {
  stop("Install the required R packages before rendering: ",
       paste(missing_r_packages, collapse = ", "))
}

# Build PDF then PNG
if (!file.exists("figures/pelger_architecture.pdf") ||
    file.mtime("figures/pelger_architecture.pdf") < file.mtime("figures/pelger_architecture.tex")) {
  invisible(tinytex::latexmk("figures/pelger_architecture.tex", engine = "pdflatex", clean = TRUE))
}
if (!file.exists("figures/pelger_architecture.png") ||
    file.mtime("figures/pelger_architecture.png") < file.mtime("figures/pelger_architecture.pdf")) {
  bmp <- pdftools::pdf_render_page("figures/pelger_architecture.pdf", page = 1, dpi = 220)
  png::writePNG(bmp, "figures/pelger_architecture.png")
}

paths <- if (knitr::is_html_output()) "figures/pelger_architecture.png" else "figures/pelger_architecture.pdf"

knitr::include_graphics(paths)
```

This architecture allows the model to learn a state-dependent SDF: the discount factor is built from portfolio weights $\omega$ that depend on both firm-level and macro information,

$$
M_{t+1} = 1 - \sum_{i} \omega\big(\text{characteristics}_{i,t},\, h_t\big)\, R^e_{i,t+1},
$$

where the weight function $\omega$ is a feed-forward network and $h_t$ is the LSTM's macro state.

To see identification within this restricted class, collect the excess returns in a vector $R^e_{t+1}$ and write $M_{t+1}=1-\omega_t^{\top}R^e_{t+1}$. The conditional Euler equation implies
$$
\mathbb{E}_t[R^e_{t+1}R_{t+1}^{e\top}]\,\omega_t
=\mathbb{E}_t[R^e_{t+1}].
$$
If the conditional second-moment matrix is nonsingular, this equation identifies $\omega_t$ within the affine tradable class. It does not identify a unique unrestricted SDF — in an incomplete market many pricing kernels satisfy the same Euler equations — and the neural-network parameters representing the same weight function need not themselves be unique. The adversarial, regularized empirical criterion selects a solution within the chosen function classes and instrument functions. Exact no-arbitrage also requires a strictly positive SDF; if strict positivity is not imposed, the fitted object is best understood as an empirical pricing-kernel candidate rather than as proof of an everywhere-positive SDF.

**Adversarial instruments.** The moment condition must hold for *all* conditioning functions $g$, which no finite procedure can impose directly. The authors' solution is a Generative Adversarial Network (GAN) framework: a second network chooses $g$ — in effect, the test assets — to make the pricing errors of the current SDF as large as possible, while the SDF network is trained to shrink them. The adversary thus supplies the hardest-to-price moments in the conditional GMM problem; it is an estimation device for the moment restrictions, not an adjustment to a prediction loss. We do not cover GANs further in this book.

**Empirical results.** The headline numbers, all from @ChenPelgerZhu2024, summarize what this combination of economic structure and neural-network flexibility delivers:

-   **Performance:** The full model (GAN-SDF) achieves an annual out-of-sample Sharpe ratio of approximately 2.6, compared with about 1.7 for linear models and 0.8 for the Fama-French 5-factor model.
-   **Explanatory Power:** The model explains over 90% of the cross-sectional variation in returns on 46 well-known anomaly portfolios.
-   **Main lesson:** In the paper's sample and benchmark design, combining economic structure (no-arbitrage moment conditions, a time-varying macro state) with machine-learning flexibility (LSTMs, adversarial estimation) performs better than the linear and prediction-only machine-learning alternatives considered. The comparison suggests that the economic restrictions do substantive work in the estimation objective.

## From Recurrent States to Attention

The LSTM cell state $C_t$ is a fixed-dimensional vector that summarizes the relevant past at each step. Its width $H$ is chosen once, before training, and does not change with the length of the input sequence. Everything a plain LSTM uses at step $t$ to forecast must pass through this bottleneck.

This design is efficient, but it has a structural cost. If the predictively relevant past at step $t$ requires distinguishing fine information from many earlier positions, an $H$-dimensional vector has to compress all of it into the same width. For moderately long sequences or for tasks where position-specific information matters — a monetary-policy statement whose meaning depends on whether a given clause appears in paragraph two or paragraph eight, for instance — the bottleneck can become binding: it binds when the predictively relevant past does not admit an adequate $H$-dimensional summary.

One of the later chapters, [Foundation Models for Economic Text and Expectations](foundation_models.qmd#sec-foundation-models), introduces a different construction. The transition from recurrent states to attention is the key structural change we emphasize in moving from the RNN/LSTM chapters to the foundation-model chapter. Attention also changes positional representation, parallel computation, and training design. We return to it formally in [From Words to Representations](foundation_models.qmd#sec-fm-representations).

**Where the implementation lives.** As in the [RNN chapter](recurrent_neural_networks.qmd), we have not fit an LSTM in code here. The [empirical chapter on networks for time series](nn_example.qmd) trains an LSTM on realized-volatility data with the complete Python pipeline — sequence construction, chronological splits, training-only scaling, tuning, and evaluation against a heterogeneous autoregressive (HAR) benchmark.

## Summary

:::{.callout-important}
## Key Takeaways
1. LSTMs modify the plain RNN state update to reduce the vanishing-gradient problem.
2. The cell state $C_t$ provides a potentially lower-attenuation path for information and gradients when the relevant forget gates remain close to one.
3. Forget, input, and output gates learn what to retain, what to add, and what to reveal at each time step.
4. The improvement comes at a cost: an LSTM has several sets of weights and is more computationally expensive than a plain RNN.
5. In econometric forecasting, LSTMs are most useful when the predictive state is likely persistent and validation performance supports the added complexity.
6. In asset-pricing applications, an LSTM can be used to represent a time-varying macroeconomic state that conditions the cross-sectional pricing function.
:::

:::{.callout-warning}
## Common Pitfalls
- Thinking that LSTMs eliminate all long-run dependence problems; they mitigate vanishing gradients, but still require enough data and careful validation.
- Interpreting the LSTM cell state as a directly observed economic state rather than a learned predictive representation.
- Using an LSTM when a lagged feed-forward network or simpler time-series model already performs well out of sample.
- Forgetting that the gates add parameters, making overfitting a serious concern in small macroeconomic samples.
- Evaluating LSTM forecasts with random splits that violate the forecast-origin information set.
:::

## Exercises

:::{.callout-note}
## Exercise 7.1: Manual LSTM Forward Pass

Consider an LSTM cell with a single unit processing one time step. You will compute all gate values, cell state, and hidden state step by step.

**Given Parameters:**

- Forget gate: $W_f = [0.3, 0.2]$, $b_f = 0.1$
- Input gate: $W_i = [0.4, 0.1]$, $b_i = 0.0$
- Output gate: $W_o = [0.2, 0.5]$, $b_o = 0.2$
- Candidate value: $W_c = [0.6, -0.3]$, $b_c = 0.0$

**Initial Conditions:**
- Current input: $x_1 = 0.5$
- Previous hidden state: $h_0 = 0.3$
- Previous cell state: $C_0 = 0.4$

**Parts:**

1. **Compute the forward pass.** Compute, in order, the forget gate $f_1$, the input gate $i_1$, the candidate value $\tilde{C}_1$, the updated cell state $C_1$, the output gate $o_1$, and the hidden state $h_1$.
2. **Interpret the update.** Decompose $C_1$ into its two terms, $f_1 \odot C_0$ and $i_1 \odot \tilde{C}_1$. Which term dominates, and what does this say about how much the cell relies on old memory versus new information at this step? Then solve for the input-gate value $i^\ast$ that would make the two terms equal, holding $\tilde{C}_1$ fixed, and explain what the result reveals about what a gate can and cannot do.
3. **Translate a gate value into a retention horizon.** Holding the gates and candidate fixed, derive $\partial C_1/\partial C_0$. Now suppose the same forget-gate value $f=0.572$ applies for $d$ consecutive steps. Derive the retained fraction of an initial perturbation after $d$ steps and solve for its half-life $d_{1/2}$, defined by $f^{d_{1/2}}=1/2$. Use $\log(1/2)\approx-0.693$ and $\log(0.572)\approx-0.559$. Explain why this calculation alone does not establish that a fitted LSTM has learned economically useful long memory.

**Note:** Use sigmoid $\sigma(z) = \frac{1}{1+e^{-z}}$ for gates and $\tanh$ for candidate values and final hidden state computation. Use the approximations $\sigma(0.29) \approx 0.572$, $\sigma(0.17) \approx 0.542$, $\sigma(0.51) \approx 0.625$, $\tanh(0.03) \approx 0.030$, and $\tanh(0.245) \approx 0.240$.

*Exam-level exercise: Part 1 is mechanical scaffolding; Part 2 interprets the competing state-update channels; Part 3 connects local sensitivity to a retention horizon and its limits.*
:::

:::{.callout-warning collapse="true"}
## Hint for Part 1
Concatenate the previous hidden state and current input: $[h_{t-1}, x_t] = [h_0, x_1] = [0.3, 0.5]$. Use this same concatenated vector in all four gate or candidate computations in Part 1.
:::

:::{.callout-tip collapse="true"}
## Solution

The LSTM equations are:
$$
\begin{aligned}
f_t &= \sigma(W_f \cdot [h_{t-1}, x_t] + b_f) \quad \text{(forget gate)} \\
i_t &= \sigma(W_i \cdot [h_{t-1}, x_t] + b_i) \quad \text{(input gate)} \\
\tilde{C}_t &= \tanh(W_c \cdot [h_{t-1}, x_t] + b_c) \quad \text{(candidate value)} \\
C_t &= f_t \odot C_{t-1} + i_t \odot \tilde{C}_t \quad \text{(cell state)} \\
o_t &= \sigma(W_o \cdot [h_{t-1}, x_t] + b_o) \quad \text{(output gate)} \\
h_t &= o_t \odot \tanh(C_t) \quad \text{(hidden state)}
\end{aligned}
$$

where $\sigma(z) = \frac{1}{1+e^{-z}}$ is the sigmoid function.

**Given:** $x_1 = 0.5$, $h_0 = 0.3$, $C_0 = 0.4$

The concatenated input vector is $[h_0, x_1] = [0.3, 0.5]$.

**Part 1: Forward pass**

*Forget gate.*
$$
\begin{aligned}
f_1 &= \sigma(W_f \cdot [0.3, 0.5] + b_f) \\
&= \sigma(0.3 \cdot 0.3 + 0.2 \cdot 0.5 + 0.1) \\
&= \sigma(0.09 + 0.1 + 0.1) \\
&= \sigma(0.29) \approx 0.572
\end{aligned}
$$

*Input gate.*
$$
\begin{aligned}
i_1 &= \sigma(W_i \cdot [0.3, 0.5] + b_i) \\
&= \sigma(0.4 \cdot 0.3 + 0.1 \cdot 0.5 + 0.0) \\
&= \sigma(0.12 + 0.05) \\
&= \sigma(0.17) \approx 0.542
\end{aligned}
$$

*Candidate value.*
$$
\begin{aligned}
\tilde{C}_1 &= \tanh(W_c \cdot [0.3, 0.5] + b_c) \\
&= \tanh(0.6 \cdot 0.3 + (-0.3) \cdot 0.5 + 0.0) \\
&= \tanh(0.18 - 0.15) \\
&= \tanh(0.03) \approx 0.030
\end{aligned}
$$

*Cell state update.*
$$
\begin{aligned}
C_1 &= f_1 \odot C_0 + i_1 \odot \tilde{C}_1 \\
&= 0.572 \cdot 0.4 + 0.542 \cdot 0.030 \\
&= 0.229 + 0.016 \\
&= 0.245
\end{aligned}
$$

*Output gate.*
$$
\begin{aligned}
o_1 &= \sigma(W_o \cdot [0.3, 0.5] + b_o) \\
&= \sigma(0.2 \cdot 0.3 + 0.5 \cdot 0.5 + 0.2) \\
&= \sigma(0.06 + 0.25 + 0.2) \\
&= \sigma(0.51) \approx 0.625
\end{aligned}
$$

*Hidden state.*
$$
\begin{aligned}
h_1 &= o_1 \odot \tanh(C_1) \\
&= 0.625 \cdot \tanh(0.245) \\
&= 0.625 \cdot 0.240 \\
&\approx 0.150
\end{aligned}
$$

**Final Results for time step 1:**

- Forget gate: $f_1 \approx 0.572$
- Input gate: $i_1 \approx 0.542$
- Candidate value: $\tilde{C}_1 \approx 0.030$
- Cell state: $C_1 \approx 0.245$
- Output gate: $o_1 \approx 0.625$
- Hidden state: $h_1 \approx 0.150$


**Part 2: Old memory versus new information**

The two terms of the update are $f_1 \odot C_0 = 0.572 \times 0.4 \approx 0.229$ and $i_1 \odot \tilde{C}_1 = 0.542 \times 0.030 \approx 0.016$. The memory term dominates by an order of magnitude: at this step the cell mostly carries forward old information and adds little from the current input. The reason is not the input gate — it is moderately open ($i_1 \approx 0.542$) — but the candidate itself: its pre-activation is only $0.03$, and since $\tanh(z) \approx z$ near zero, the candidate $\tilde{C}_1 \approx 0.030$ is itself close to zero, so there is little new content to add, however open the gate. The moderately open forget gate ($f_1 \approx 0.572$) passes on roughly half of the previous cell state.

For the two terms to be equal with $f_1$, $C_0$, and $\tilde{C}_1$ held fixed, the input gate would need $i^\ast = 0.229 / 0.030 \approx 7.6$ — far outside the sigmoid's range $(0,1)$. No admissible input-gate value can make the new information dominate under this counterfactual. A gate rescales its channel by a factor between 0 and 1, so it can attenuate a signal but never amplify one. With the current carry-over term fixed, dominance would require a larger candidate magnitude. In the unrestricted cell update, a smaller forget-gate contribution $f_1C_0$ is the other route by which new information can dominate.

**Part 3: Retention horizon**

Holding the gates and candidate fixed gives
$$
\frac{\partial C_1}{\partial C_0}=f_1=0.572.
$$
If the same forget gate applies for $d$ steps, the retained fraction is the product $f^d=0.572^d$. The half-life therefore solves
$$
d_{1/2}=\frac{\log(1/2)}{\log(0.572)}
\approx\frac{-0.693}{-0.559}\approx1.24\text{ steps}.
$$
This coordinate would attenuate an initial perturbation by one half after only about 1.24 steps under the constant-gate approximation, so the computed value does not indicate long retention. More generally, one realized gate value does not describe the time-varying products along other observations, and even a near-one gate says nothing about whether the retained state improves forecasts; Exercise 7.2 takes up the validation evidence that such a claim requires.

**Further remarks:**

1. **Selective Output**: The output gate ($o_1 = 0.625$) filters what portion of the cell state becomes the hidden state, demonstrating how LSTMs can store information internally without immediately exposing it.

2. **Computational Complexity**: Even for a single time step with one unit, LSTMs require computing four different weight-input combinations, highlighting why they are computationally more expensive than plain RNNs.

:::

:::{.callout-note}
## Exercise 7.2: The Cell-State Gradient Highway

Consider a scalar LSTM cell with state update
$$
C_k = f_k\, C_{k-1} + i_k\, \tilde{C}_k,
$$
where the forget gate $f_k \in (0,1)$, the input gate $i_k \in (0,1)$, and the candidate $\tilde{C}_k \in (-1,1)$ are computed from $(h_{k-1}, x_k)$ as in the chapter. In Parts 1 and 2, use the **gates-fixed approximation** from the chapter: treat $f_k$, $i_k$, and $\tilde{C}_k$ as fixed numbers that do not vary with the earlier cell state.

1. Derive
$$
\frac{\partial C_T}{\partial C_t} = \prod_{k=t+1}^{T} f_k
$$
under the gates-fixed approximation, starting from the one-step derivative $\partial C_k / \partial C_{k-1}$.

2. Compare this with the one-dimensional RNN sensitivity from [Exercise 6.1](recurrent_neural_networks.qmd#rnn-exercise-6-1), $\partial h_T/\partial h_t = \prod_{k=t+1}^{T} W_{hh} \tanh'(z_k)$, whose magnitude is bounded by $|W_{hh}|^{T-t}$. Suppose $W_{hh} = 0.8$ (as in [Exercise 6.2](recurrent_neural_networks.qmd#rnn-exercise-6-2)) while the LSTM has learned $f_k = 0.99$ for every $k$. For a horizon of $T-t = 30$ steps, evaluate the RNN upper bound and the LSTM cell-path magnitude (use $0.8^{30} \approx 0.0012$ and $0.99^{30} \approx 0.74$) and explain which two structural features of the cell-state update produce the difference: what replaces the repeated multiplication by $W_{hh}$, and why does no activation derivative appear in the LSTM product?

3. Part 1 holds the gates fixed. Identify the channels through which gradients flow in a *full* backward pass that this approximation ignores, and explain why the claim "LSTMs eliminate vanishing gradients" therefore does not follow from Part 1. Then explain why this limitation makes an out-of-sample comparison against a lagged feed-forward or heterogeneous autoregressive (HAR) benchmark necessary before attributing forecast gains to learned long memory.

*Exam level. Part 1 is a short chain-rule derivation; Part 2 rewards understanding of what structurally differs between the two recursions; Part 3 tests whether students can locate the limits of the gates-fixed result.*
:::

:::{.callout-warning collapse="true"}
## Hint for Part 3
Ask which quantities at step $k+1$ are computed from $h_k$, and what $h_k$ itself is computed from.
:::

:::{.callout-tip collapse="true"}
## Solution

**Part 1: The gates-fixed derivation**

Under the gates-fixed approximation, the only dependence of $C_k$ on $C_{k-1}$ is the explicit linear term:
$$
\frac{\partial C_k}{\partial C_{k-1}} = f_k,
$$
since $i_k \tilde{C}_k$ is treated as a constant. By the chain rule along the cell-state path,
$$
\frac{\partial C_T}{\partial C_t}
= \prod_{k=t+1}^{T} \frac{\partial C_k}{\partial C_{k-1}}
= \prod_{k=t+1}^{T} f_k.
$$

**Part 2: Comparison with the RNN recursion**

For the RNN, $|\partial h_T/\partial h_t| \le |W_{hh}|^{T-t}$ because $|\tanh'(z)| \le 1$; with $W_{hh} = 0.8$ and $T-t = 30$, the bound is $0.8^{30} \approx 0.0012$ — the gradient signal from step $t$ is essentially extinguished. For the LSTM with $f_k = 0.99$, the cell-state path carries $0.99^{30} \approx 0.74$ — about three-quarters of the signal survives thirty steps.

Two structural features produce the difference. First, $C_{k-1}$ enters the cell-state update linearly, multiplied only by a learned, input-dependent forget gate in $(0,1)$ — a separate, adaptive retention channel. A scalar RNN recurrent weight can also be set near one, but the same weight must simultaneously implement the state dynamics, so it cannot be tuned for retention alone. Second, the direct cell-state path bypasses the activation function: $C_{k-1}$ is not passed through a $\tanh$ on its way into $C_k$, so no activation derivative enters this product, whereas every RNN step contributes a factor $\tanh'(z_k)\le 1$ that attenuates further whenever the unit saturates.

**Part 3: What the gates-fixed approximation ignores**

In a full backward pass, gradients also flow through channels that Part 1 freezes: the hidden state $h_k = o_k \odot \tanh(C_k)$ feeds the gate computations at step $k+1$, so $f_k$, $i_k$, $o_k$, and $\tilde{C}_k$ all depend on earlier cell states through $h_{k-1}$. Those channels involve products of sigmoid and tanh derivatives with weight matrices and can attenuate exactly as in a plain RNN. In addition, the sigmoid never reaches 1 exactly, so $f_k < 1$ strictly and even the highway attenuates — just controllably slowly. Part 1 therefore establishes that *one component* of the backward pass has a weight-free, learnable path: a mitigation of the vanishing-gradient problem, not its elimination.

Practical implication: do not assume the architecture alone guarantees that long-run dependence is learned. Long-horizon LSTM forecasts still require enough data and out-of-sample validation evidence — for instance, a comparison against a lagged feed-forward or HAR-type benchmark, as in the empirical chapter — before attributing performance to learned long memory.
:::

:::: {.content-visible when-format="html"}
## References

::: {#refs}
:::

::::
